Realizing temporal graphs from fastest travel times

### 2023-03-22 Response.txt

We would like to thank the reviewers for their work and comments.

ANSWER TO REVIEWER 1:
The main issue the reviewer had was with the motivation of the problem. It is remarked that the idea of extending the problem from a static case to the temporal is not interesting enough on its own, while this is one of the main approaches in the field of temporal graphs - how to (in the most natural way) extend the problems from static graphs to temporal, and then study the complexity of derived problem(s).
We introduced the idea of periodic edges as this makes two vertices trivially connected, which makes the argument of finding a temporal path of fastest duration more interesting.
We believe the idea of realizing fastest temporal paths is the most natural as it is easy to calculate these values and therefore check if temporal graph is really 'good' (in the opposite to proposed "minimization of the maximum/average duration under some constraints").

Since we proved that the problem is NP-hard in general and polynomial-time solvable on trees, the next natural question is how can the tree-like structure help us with solving the problem exactly. We show that FPT is possible only in the case when parameter is FES (which we agree, it is not a 'good' parameter in practice), but we show that this is the smallest parameter that still gives us an FPT algorithm. In the case of parameterizing the problem already with the FVS we show that the problem is W[1]-hard. Therefore all other 'smaller' parameters, cannot give us an FPT result.

We thank the reviewer for suggesting the paper "Dynamic networks: models and algorithms" by Kuhn and Oshman, 
we will cite it in our final version.

ANSWER TO REVIEWER 2:
Definition 13 defines segments as paths between two vertices of interest. There cannot exist more than two segments among the same pair of vertices of interest (where one is just an edge). As two vertices of interest are the endpoints of one edge (from FES), therefore one segment is an edge and the other is the rest of the cycle. In the case when vertices of interest have degree > 2 then there is a unique segment among each pair of them.

ANSWER TO REVIEWER 3:
Each connector gadget contributes 2 vertices to the FVS. We will add the missing edges to the figure.
We will also define the MULTICOLORED CLIQUE problem, as suggested.

### 2023-03-24 Response.txt

We would like to thank the reviewers for their work and comments.

ANSWER TO REVIEWER 1:
Regarding the movitation of periodicity: We want to point out that, since every edge having the same period is clearly a special case of every edge having *some* period, our hardness results also hold for the more general/realistic setting that the reviewer described. An even more general setting would be to allow more than one label per edge per periods (this can model different periods by taking the LCM of them as a global period). We propose this as a future research direction in the conclusion (lines 603-604). However, we believe that thoroughly understanding the fundamental case of having the same period and exactly one label per edge per period is vital before moving towards understanding more general/realistic settings.

Regarding the motivation of considering durations: We agree that from the "design-perspective" many other properties to realize might be well-motivated, such as the ones proposed by the reviewer. However, from the "verification-perspective" (cf. lines 28-31) it is well-motivated that a matrix of e.g. measured duration-values is given, and we are asked to reconstruct the temporal network displaying exactly those durations.

Regarding the motivation of studying parameterized complexity: This is a canonical and well-established way to approach NP-hard problems when exact algorithms are desired. We did not study approximation because our problem does not have a canonical formulation as an optimization problem. What would be the objective function? If we want to get the durations as close as possible to the ones prescribed in the matrix, it opens new questions such as e.g. "How many edges are we allowed to put into the graph?". Since this is a study of a new problem, we believe it is natural to initially focus on exact algorithms.

Since we identified a restricted setting (where the underlying graph is a forest) in which our problem is solvable in polynomial time, it is natural to explore how this algorithm generalizes. This polynomial-time solvable special case of trees gives a strong (theoretical) motivation to study parameters that measure "tree-likeness". From a practical standpoint, for many real-world networks it is well-motivated to assume that they are very structured, which motivates the study of structural graph parameters. Tree-likeness is a fairly common assumption, in particular for transportation networks (see e.g. reference [19]).

Note that our main parameterized hardness result (Theorem 4) rules out the possibility of obtaining FPT algorithms for many structural graph parameters, and not only for the ones that measure tree-likeness. So the question rather becomes "Which structural properties of the underlying graph allow for efficient algorithms?". Note that e.g. the arboricity can be upper-bounded by a function of the feedback vertex number, and hence we cannot expect to obtain fixed-parameter tractability for this parameter. Note that also parameters that can be small in dense graphs, such as the cliquewidth or the vertex-deletion-distance to chordal graphs, can be upper-bounded by a function of the feedback vertex number. Hence, we do have a strong (theoretical) justification for obtaining fixed-parameter tractability with the feedback edge number (although it is a rather large parameter), because we can rule out fixed-parameter tractability for most smaller parameters (as discussed e.g. in lines 363-364). We indeed leave the parameterized complexity wrt. pathwidth open, and in the conclusion of the full version we suggest studying this (see appendix lines 2198-2199).

We thank the reviewer for pointing us towards "dynamic graphs" and we will add a discussion about this to the related work section. However, this is a related but *different* mathematical model. Note that we explicitly construct a (periodic) temporal graph, that is, the mathematical object contains all information about how the graph changes (deterministically). In dynamic graphs, as described in the work "Dynamic networks: models and algorithms" by Kuhn and Oshman, the assumption is that the changes are adversarial or random (while obeying some constraints such as connectivity) and not necessarily known in advance.

ANSWER TO REVIEWER 2:
We made great effort to provide intuitive explanations of the main ideas and techniques we use, as well as some technical details that convey an intuition about the remaining steps necessary to complete the proofs (which are given in the appendix). We believe that our work is relevant to a larger audience as it initiates the research of a well-motivated new problem setting and generates several future research directions. We see much potential for positive results.

Definition 13 defines segments as paths between two vertices of interest. Indeed, these path segments should be non-trivial (i.e. they must have at least two edges), we will clarify this in teh final version. If there were two (non-trivial) segments between two vertices of interest, then there would be a cycle which contains a feedback edge. It follows that one of the segments must contain a further vertex of interest, which is not allowed by the definition.

ANSWER TO REVIEWER 3:
Each connector gadget contributes 2 vertices to the FVS, since many other vertices are connected to \hat{v}_1 and \hat{v}_2. The edges are omitted from the picture to avoid cluttering. We will add indications of the missing edges to the figure and clarify this in the figure caption.

We provide a definition of the problem MULTICOLORED CLIQUE in lines 187-188.

### 2023-04-12 New Paper/Realizing TG - extended.tex
Added algorithm for D upper bounds for paths - as discussed with George and Paul on 11th of April

%% One label per edge, period Delta, upper-bounded realization
\newcommand{\kDeltaUpperBoundLong}{\textsc{Simple $\Delta$-periodic Upper-Bounded Temporal Graph Realization}}
%\newcommand{\deltaUpperBound}{}

@@

\problemdef{\kDeltaUpperBoundLong}
{A static graph $G=(V,E)$ with $n$ vertices, $n \times n$ matrix $D$ of positive integers and a bijective mapping $\phi$ from vertices $V$ of $G$ to the rows/columns of the matrix $D$.}
{Does there exist a $\Delta$-periodic labeling $\lambda: E \rightarrow \{1,2,\ldots,\Delta\}$ such that, 
	for every $i,j$, the duration of the fastest temporal path from $v_i$ to $v_j$ in the $\Delta$-periodic temporal graph $(G,\lambda,\Delta)$ is \emph{at most} $D_{\phi(i),\phi(j)}$.}

@@

\section{Upper bounds in the matrix D}

@@

\problemdef{\kDeltaUpperBoundLong}
{An $n \times n$ matrix $D$ of positive integers and an integer $k\in \mathbb{N}$.}
{Does there exist a graph $G=(V,E)$ with vertices $\{v_1,\ldots,v_{n}\}$ with $n+k-1$ edges 
	and a $\Delta$-periodic labeling $\lambda: E \rightarrow \{1,2,\ldots,\Delta\}$ such that, 
	for every $i,j$, the duration of the fastest temporal path from $v_i$ to $v_j$ in the $\Delta$-periodic temporal graph $(G,\lambda,\Delta)$ is \emph{at most} $D_{i,j}$.}

>>>>

%% One label per edge, period Delta, upper-bounded realization
\newcommand{\DeltaUpperBoundLong}{\textsc{Simple $\Delta$-periodic Upper-Bounded Temporal Graph Realization}}
\newcommand{\deltaUpperBound}{\textsc{Simple Upper-Bounded TGR}}

@@

\problemdef{\DeltaUpperBoundLong}
{A static graph $G=(V,E)$ with $n$ vertices, $n \times n$ matrix $D$ of positive integers and a bijective mapping $\phi$ from vertices $V$ of $G$ to the rows/columns of the matrix $D$.}
{Does there exist a $\Delta$-periodic labeling $\lambda: E \rightarrow \{1,2,\ldots,\Delta\}$ such that, 
	for every $i,j$, the duration of the fastest temporal path from $v_i$ to $v_j$ in the $\Delta$-periodic temporal graph $(G,\lambda,\Delta)$ is \emph{at most} $D_{\phi(i),\phi(j)}$.}

@@

\section{Upper bounds in the matrix D}
We start by investigating the hardness of the problem on paths.
We first reduce the input matrix $D$, by performing a series of modifications.

Denote with $P = (v_1, v_2, \dots, v_n)$ a path on $n$ vertices,
and let $D$ be the duration matrix with upper bounds.
For any two vertices $v_i, v_j$ from $P$ we get that:
\begin{itemize}
	\item if the value of $D_{i,j} < D_{i,k}$, where $k \in [1+1,j-1]$
	we set $D_{i,k} = D_{i,j}$.
\end{itemize}
%
By the definition of temporal paths, it follows that $d(i,j) = d(i,k) + d(k-1, j) - 1$, where $i,j \in [n]$, with $i<j$ and $k \in [i+1,j-1]$.
Therefore,
\begin{itemize}
	\item if $D_{i,j} > D_{i,k} + D_{k-1,j} -1$ we set the value of $D_{i,j}$ to $D_{i,k} + D_{k-1,j} -1$. 
\end{itemize}
%
And lastly, to any $i$-th row of the matrix $D$, we make the following modifications:
\begin{itemize}
	\item if there are two entries $j< k$ with the same values, \ie $D_{i,j} = D_{i,k}$ where both $j,k$ are bigger than $i$ (resp.~smaller than $i$),
	then we set $D_{i,j} = D_{i,k} - (k-j)$ (resp.~ $D_{i,k} = D_{i,j} - (k-j)$).
\end{itemize}
We iteratively repeat all of the above steps and produce a reduce matrix $D'$.
It is easy to see that if there exists a labeling of $P$ for $D'$, then this labeling satisfies also $D$.

\begin{theorem}
	\deltaUpperBound\ can be solved in polynomial time on paths.
\end{theorem}

\begin{proof}[Proof idea:]
	Let $P = (v_1, v_2, \dots, v_n)$ be a path on $n$ vertices.
	For the sake of simplicity we denote with $a_i$ label of edge $e_i =v_iv_{i+1}$, for all $i \in [n-1]$.
	By the definition, for any two vertices $v_i,v_j \in P$, we have the following inequality:
	\begin{equation*}
		(a_{i+1} - a_i)_\Delta + (a_{i+2} - a_{i+1})_\Delta + \cdots + (a_{j-1} - a_{j-2})_\Delta + 1
		\leq D_{i,j}.
	\end{equation*}
	For the sake of simplicity we refer to the inequality that a temporal path from $v_i$ to $v_j$ has to respect (\ie~the above inequality) 
	as $I_{i,j}$.
	Note that we can rewrite the inequality $I_{i,j}$ also without the modulo $\Delta$ part as:
	\begin{equation*}
		k_{i+1,i} \Delta + (a_{i+1} - a_i)+ k_{i+2,i+1} \Delta + (a_{i+2} - a_{i+1}) + \cdots + k_{j-1,j-2} \Delta + (a_{j-1} - a_{j-2})+ 1
		\leq D_{i,j},
	\end{equation*}
	where all $k_{i,i+1} \in \{0,1\}$.

	W.l.o.g. we can assume also that $a_i \neq a_{i+1}$ for all $i \in [n-1]$,
	this follows from the fact that the duration among two vertices $v_i, v_{i+2}$ on distance $2$ in $P$ equals to $\Delta + 1$ when the two consecutive labels are the same, but if the labels are not the same, the duration can be at most $\Delta$, which still satisfies the condition from before.
	Therefore, for any two consecutive labels $a_i$ and $a_{i+1}$ (\ie for any two vertices $v_i, v_{i+2}$ on distance $2$ in $P$), 
	it follows that 
	$k_{i, i+1} = 1 - k_{i+1,i}$, where $k_{i+1,i} \in \{0,1\}$, which makes also $k_{i,i+1} \in \{0,1\}$, as required.
	If we now sum up the two inequalities $I_{i,i+2}, I_{i+2,i}$ we get the following:
	\begin{equation*}
		(a_i-a_{i+1})_\Delta + (a_{i+1}-a_i)_\Delta +1 +1 = \Delta + 2.
	\end{equation*}
	It follows that for any two vertices $v_i, v_j$ in $P$, that are on distance $k$ apart, 
	we get that $d(i,j) + d(j,i) = (k-1) \Delta + 2$.

	To satisfy the inequalities $I_{1,n}$ and $I_{n,1}$ we have to correctly determine which of $k_{i+1,i}$ have value $1$, and consequently set $k_{i,i+1} = 0$,
	and vice versa.

	From the value $D_{1,n}$ (resp.~$D_{n,1}$) we can determine how many of $k_{i+1,i}$ (resp.$k_{i,i+1}$) can be non-zero. Suppose that there are $m$ non-zero ones (resp.~$n-m-1$).
	Then to determine which $m$ of the $k_{i,i+1}$ are non-zero we look at every two vertices $v_i, v_{i+2}$ at the distance $2$ in $P$ 
	and based on the inequalities $I_{i,i+2}, I_{i+2,i}$ determine if $k_{i+1,i}$ has a value $1$ or $0$,
	i.e.~if $e_{i+1}$ is bigger or smaller than $e_i$.
	We start determining the values $k_{i+1,i}$ from ``left to right'' (i.e~first we determine $k_{2,1}$, then $k_{3,2}$, \dots, and at last $k_{i-1,i-2}$).
	Now that the relations of every two consecutive labels of $P$ are determined, 
	we select the values for each label, such that the inequalities among them hold.
	\begin{claim}
		The above procedure produces a labeling of $P$ that satisfies durations from $D$.
	\end{claim}
\end{proof}

@@

\problemdef{\DeltaUpperBoundLong}
{An $n \times n$ matrix $D$ of positive integers and an integer $k\in \mathbb{N}$.}
{Does there exist a graph $G=(V,E)$ with vertices $\{v_1,\ldots,v_{n}\}$ with $n+k-1$ edges 
	and a $\Delta$-periodic labeling $\lambda: E \rightarrow \{1,2,\ldots,\Delta\}$ such that, 
	for every $i,j$, the duration of the fastest temporal path from $v_i$ to $v_j$ in the $\Delta$-periodic temporal graph $(G,\lambda,\Delta)$ is \emph{at most} $D_{i,j}$.}
